\section{Silent area}
Visitors who need an acoustic break or want to deepen their La La Lab experience are welcome to the silent area in the exhibition space. Here you can discover the following:

You can sit in our library and browse books and articles on the field of Mathematics and Music, a rich domain of current research with dozens of modern books on the subject. The domain got a major impulse with the Society for Mathematics and Computation in Music, founded in 2006. You can find here its Journal of Mathematics and Music, and the proceedings of its conferences. Also a rich selection of music-related articles from the Bridges conferences on art and mathematics, and some other selected articles and books on the subject.

Did you always want to know how certain sequences of numbers sound when transformed into sounds? Choose the ``The Sound of Sequences'' app and find out more.

Use the ``Note compass'' to explore diatonic scales, to understand why the piano has white and black keys, and to hear the difference of styles they convey.

Move through the ``Space of Pentatonic Scales'' and create your own impressionistic style music by playing pieces by classical composers with only five notes.

\begin{sectcredits}
\item[With contributions from:] IMAGINARY, Neil Sloan, Jürgen Richter-Gebert, Aaron Montag, Thomas Noll, various researchers and musicians.
\item[Special thanks to:] OEIS Foundation, Springer Publishing House and Taylor \& Francis Publishing House.
\end{sectcredits}
